---
title: "EPSY 1261 Homework: Box-plots"
author: "YOUR NAME GOES HERE"
date: "June 22, 2016"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
require(natality2014)
Birth_sample <- Natality_2014_1k

```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
```

In this homework assignment you will choose a quantitative variable and at least one categorical variable from the Natality_2014_100k package to make interesting box-plots. Start by writing an engaging question, then using the code from the Class Activity: Box-plot make graphs and summary statistics. You may need to change variables and questions as you look for interesting relationships. When you have your final research question, graphs, and summary statistics write about your relationship. 

You will choose your variables from the Natality_2014_100k package, but you will use the `Birth_sample.Rmd` data frame to make your graphs and statistics. Remember `Birth_sample.Rmd` is just a random sample of 1000 observations from the Natality_2014_100k package.

Recall that data sets for most of the tables, examples and exercises are available online.  See examples and hints below. 

Steps to proceed:

1) download the files `Birth_sample.Rmd` and `HWBoxplot.Rmd` from Moodle

2) upload the files to the RStudio server

3) run "Knit PDF" and read the directions for this homework assignment

4) then go back to the `HWBoxplot.Rmd` file and add your code and comments

5) "Knit PDF" and double check your work

6) Upload your "Knit PDF" to the Moodle homework assignment.

devtools::install_github("dtkaplan/natality2014")

## PROBLEM TO COMPLETE:

### 1. To choose the variables you want to work with you may want to do the following.

To view the variable names in the Natality_2014_100k package run the following code: names(Birth_sample)

```{r,fig.height=3.2}

```
To view the first six rows of the data in `Birth_sample.Rmd` run the following code: head(Birth_sample)

```{r,fig.height=3.2}

```
To better understand the variable names do the following
1) Go to the Packages Tab and click the package `natality_2014_100k`
2) Then click on `Natality_2014_100k`
3) You should see the description of the data and below that all the list of variables in the data frame with their description


### 2. Write your research question, make sure to explain which variable you are investigating.



### 3. Write your code to make the box-plots and summary statistics 
Run the code. Your assignment will include code, visualizations, statistics, and written interpretation.
```{r, fig.height=3.2}


```


### 4. Explain your relationship referring to the graphs and statistics
Make sure to explain the variables you are using and include sample sizes, shape, center, spread and outliers. What is the 'take away' from your research?