---
title: 'A “second date" with your data in R/RStudio'
author: "Alison Link"
date: "June 20, 2016"
output: 
  html_document:
    css: styles.css
---

***

**Related to:** _Data Computing_, "Introduction to Data Graphics", Ch. 5

***

```{r, echo=FALSE}
data <- read.csv("Benchmarking_Results__CY_2012_Public_Buildings.csv", header=TRUE, na.strings=c("N/A", "Not Available"))
```

## Check out some "profile pictures" of your data

(In real dating, you probably would have done this earlier on in the process, but for data dating, we'll save the fun stuff for last.)


## Look at a single variable

### Boxplot

[graphic with parts of boxplot labeled]

If you remember the `summary()` function we looked at earlier, a boxplot is essentially a visual representation of those values--minimum, median, maximum, etc.  Here's how you create a boxplot in R:

```{r}
boxplot(data$ENERGY_STAR_score)
```

We can also group our boxplot, to examine how values vary between groups.  For example, we have a variable called "Prop\_Type" in our dataset that describes whether the building is a school, office, parking lot, etc.  Maybe we're wondering how the Energy Star rating tends to vary based on property type.  Let's group buildings with the same "Prop\_Type" together, then, and see how that changes our boxplot:

```{r}
par(las=2, mar=c(10, 4, 2, 2))
boxplot(data$ENERGY_STAR_score ~ data$Prop_Type)
```

_Wait a minute...that boxplot looks kind of empty!_ you may be saying.  And you're right: there are clearly values on the chart that have no associated boxplot (for example: "Convention Center").  This is generally because there are not enough values in those group to be able to calculate a meaningful boxplot--there's only one Convention Center in Minneapolis, after all!  This is a good lesson: when grouping based on a variable like this, it's always good to check how many values fall into each group before plotting or doing further analysis.  If you don't have enough observations in a particular group, you will need to account for that during your analysis.  You can see how many obeservations fall in each "Prop_Type" group using the `table()` function:

```{r}
table(data$Prop_Type)
```

### Activity A {.activity}

Use the `table()` function to find out: how many parking lots do we have in this dataset?  Parking lots also don't have a boxplot associated with them in the chart above.  Why do you think that is?


### Histogram

### Barchart



## Look at two variables

### Scatterplots


## References

City of Minneapolis. (2012). _Benchmarking Results - CY 2012 Public Buildings_ [Data set]. Retrieved from: http://opendata.minneapolismn.gov/datasets/d050496ced314752bf89763139bc0cb3_0 


***

**Acknowledgements:** Thanks to Roxanne Johnson for helping to brainstorm some of the topics in this document!
